---
title: "Introduction to ImmuPop"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to ImmuPop}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

ImmuPop estimates population-level immunity from individual antibody titer
data using bootstrap sampling and MCMC-based Dirichlet estimation with
age-structured contact matrices. The package produces four key metrics:

- **Population immunity** — the effective immune proportion weighted by
  age-specific contact patterns and protection curves.
- **Relative reduction in R0** — how much transmission potential is reduced
  relative to a fully susceptible population.
- **Geometric mean titer (GMT)** — the population-weighted geometric mean
  antibody titer.
- **Proportion seropositive** — the proportion of individuals with
  detectable antibody (titer ≥ 1:10).

## Quick start

```{r quickstart}
library(ImmuPop)

# Load bundled example data
data("ImmuPop_raw_data")
head(ImmuPop_raw_data)
```

The raw data must contain at least `age` and `raw_titer` columns. It may
also include `time`, `epi` (epidemic identifier), and `baseline` columns
for longitudinal and multi-epidemic analyses.

### Step 1: Prepare data

Use `generate_data()` to bin ages into groups and compute log-transformed
titer levels:

```{r prepare}
df <- generate_data(ImmuPop_raw_data, cut_age = c(0, 18, 100))
head(df)
```

### Step 2: Define population parameters

ImmuPop requires three inputs describing the population:

```{r params}
# Age group proportions (must sum to 1, one per age group)
age_prop <- c(0.2, 0.8)

# Contact matrix between age groups (rows = contactors, cols = contactees)
contact_matrix <- matrix(c(22, 16, 28, 120), nrow = 2, byrow = TRUE)

# Protection curves: probability of protection at each titer level
# (one value per titer level, from lowest to highest)
protect_c <- c(0.1, 0.2, 0.3, 0.4, 0.5)  # children
protect_a <- c(0.1, 0.2, 0.3, 0.4, 0.5)  # adults
```

### Step 3: Estimate immunity at a single time point

```{r timepoint}
# Subset to one time point
data_t <- df[df$time == 2, ]

result <- ImmuPop_est_timepoint(
  df             = data_t,
  protect_c      = protect_c,
  protect_a      = protect_a,
  age_prop       = age_prop,
  contact_matrix = contact_matrix,
  sim_num        = 200,
  seed           = 42
)
result
```

### Step 4: Visualize results

```{r plot-forest, fig.height=4}
plot_estimates(result)
```

## Multi-epidemic baseline comparison

When data span multiple epidemics, `ImmuPop_est_baseline()` estimates
immunity at the pre-epidemic baseline for each epidemic:

```{r baseline}
df_bl <- df[df$baseline == "yes", ]

result_bl <- ImmuPop_est_baseline(
  df_baseline    = df_bl,
  protect_c      = protect_c,
  protect_a      = protect_a,
  age_prop       = age_prop,
  contact_matrix = contact_matrix,
  sim_num        = 200,
  seed           = 42
)
result_bl
```

```{r plot-baseline, fig.height=5}
plot_estimates(result_bl)
```

## Time series estimation

`ImmuPop_est_timeseries()` loops over all time points and returns a
combined data frame suitable for trend visualization:

```{r timeseries}
result_ts <- ImmuPop_est_timeseries(
  df_long        = df,
  protect_c      = protect_c,
  protect_a      = protect_a,
  age_prop       = age_prop,
  contact_matrix = contact_matrix,
  sim_num        = 200,
  seed           = 42
)
head(result_ts)
```

With many time points (>8), `plot_estimates()` automatically switches to a
line plot with shaded 95% CI ribbon:

```{r plot-timeseries, eval=FALSE}
# Requires > 8 time points for line-plot mode
plot_estimates(result_ts)
```

## Titer distribution plots

Visualize the raw titer distribution across age groups:

```{r plot-dist, fig.height=4}
plot_titer_dist(df[df$time == 2, ], main = "Titer distribution (time 2)")
```

## Titer jitter plot

Individual titers with GMT overlay and reference thresholds:

```{r plot-jitter, fig.height=4}
plot_titer_jitter(df[df$time == 2, ], main = "Time 2", seed = 1)
```

## Saving plots to PDF

All plot functions accept a `file` argument for direct PDF export:

```{r pdf-export, eval=FALSE}
plot_estimates(result, file = "forest_plot.pdf")
plot_titer_jitter(df[df$time == 1, ], file = "jitter.pdf")
```
